### 2025-11-08 Proposals/2025-NSF-EntropicGeometry/NarrativeBody.tex
Final version of project summary, significance, and bibliography formatting before talking to Alejandro

The second axiom imposes a convex structure, which represents the ability to perform statistical mixtures. Given two ensembles $\ens[a], \ens[b] \in \Ens$, we can always choose a probability $p \in [0,1]$ and select $\ens[a]$ $p$ percent of the times and $\ens[b]$ $\bar{p} = 1-p$ percent of the times. This gives the ensemble space a convex structure as $p \ens[a] + \bar{p} \ens[b] \in \Ens$ for all $\ens[a], \ens[b] \in \Ens$ and $p \in [0,1]$. This insight is already used in the literature, particularly in Generalized Probabilistic Theories (GPTs); see e.g.~Refs.~\cite{plavala2023,wilce2025generalizedprobabilitytheorynotes,DArianoBook} \textcolor{blue}{Anything else to cite?}. The main innovation in this respect is that we do not limit ourselves to finite combinations and we do not simply allow all infinite combinations (which can be shown to include physically untenable cases~\cite{baccetti2013infinite,aop-unphysHilbert}). Rather, we let the topology do the work. That is, infinite convex combinations exist only if they are the topological limit of finite convex combinations. On physical grounds, the mixing operation must be continuous in the natural topology defined by experimental verifiability, much like addition and scalar multiplication are continuous in a topological vector space, leading to the idea of a topological convex space.

@@

\noindent
\textbf{Expected outcomes and significance:}
During the three-year period of the proposed research, we expect to achieve \textcolor{blue}{and submit manuscripts on} at least the following:
\begin{itemize}
	\item Demonstrate that the Shannon entropy can be recovered as the maximum entropy increase under mixture for all ensemble spaces, and that the Shannon, Gibbs and von Neumann entropies can be recovered in the specific cases of discrete classical, continuous classical, and quantum mechanics.
	\item Show that the geometric structures of classical and quantum mechanics can be recovered from the entropic structure alone.
	\item Demonstrate that information geometry can be generalized to the ensemble space of any physical theory of finite dimensions. 
	\item Show that the entropic structure imposes constraints on the space of possible physical theories and rules out some Generalized Probabilistic Theories.
\end{itemize}
The above expected results should lead naturally to a range of additional possible directions and questions that could be addressed during or subsequent to the grant period, depending on the rate of advancement.  For example:
\begin{itemize}
	\item Identify what results from information geometry and information theory could be generalized to ensemble spaces; this would lend itself to various self-contained contributions by the grad student, undergrads, or individuals outside the project.
	\item Determine whether the convexity of the mixing entropy can be physically motivated and whether that guarantees a continuous embedding into a locally convex topological vector space.
	\item Generalize information geometry to the infinite-dimensional case; this would require a locally convex topology to be able to define variations.
	\item Determine whether the ensemble spaces can be in general proven to be metrizable and characterize the connection between the metric and the entropy.
	\item Find a suitable function space that can be used as a model for the case of classical field theories.
	\item Generalize classical Hamiltonian mechanics and the principle of stationary action to a generic ensemble space.
\end{itemize}

\textcolor{blue}{The immediate significance of the proposed research stems from the realization that, though classical information theory and quantum information theory are presented as two separate subjects in textbooks, there are a lot of commonalities that make them an interesting target for generalization.  
The broader significance lies in its approach as well as its specific potential to provide a common axiomatic foundation for \emph{all} physical theories.  Currently within physics, we typically do not know from where the various mathematical structures emerge, making it unclear in which directions they can be generalized in order to develop new theories that are physically meaningful.  Our approach instead seeks to \emph{find axioms from physical requirements}, i.e.~the axiomatics are justified from the physics.  Such an axiomatic foundation common to all physical theories would serve not only to better ground current theories and elucidate their relationships, but also to constrain potential future physical theories.}

@@

While there exists other work attempting to reconstruct quantum mechanics from physical principles, for example in Refs.~\cite{hardy2001quantumaxioms,caticha2008,dakic2009,Chiribella2010,Masanes_2011}, we are not aware of any literature attempting to relate the geometric and entropic structures of either classical or quantum mechanics, much less those of a general physical theory. Works trying to reconstruct quantum mechancis from physical premises are typically in finite dimensions, do not generalize to classical mechanics, and do not provide a mathematically rigorous path to generalize to classical or quantum field theories.

>>>>

The second axiom imposes a convex structure, which represents the ability to perform statistical mixtures. Given two ensembles $\ens[a], \ens[b] \in \Ens$, we can always choose a probability $p \in [0,1]$ and select $\ens[a]$ $p$ percent of the times and $\ens[b]$ $\bar{p} = 1-p$ percent of the times. This gives the ensemble space a convex structure as $p \ens[a] + \bar{p} \ens[b] \in \Ens$ for all $\ens[a], \ens[b] \in \Ens$ and $p \in [0,1]$. This insight is already used in the literature, particularly in Generalized Probabilistic Theories (GPTs); see e.g.~Refs.~\cite{plavala2023,wilce2025generalizedprobabilitytheorynotes,DArianoBook}. The main innovation in this respect is that we do not limit ourselves to finite combinations and we do not simply allow all infinite combinations (which can be shown to include physically untenable cases~\cite{baccetti2013infinite,aop-unphysHilbert}). Rather, we let the topology do the work. That is, infinite convex combinations exist only if they are the topological limit of finite convex combinations. On physical grounds, the mixing operation must be continuous in the natural topology defined by experimental verifiability, much like addition and scalar multiplication are continuous in a topological vector space, leading to the idea of a topological convex space.

@@

\noindent
\textbf{Expected outcomes and significance:}
During the three-year period of the proposed research, we expect to achieve and submit manuscripts on at least the following:
\begin{itemize}
	\item Demonstrate that the Shannon entropy can be recovered as the maximum entropy increase under mixture for all ensemble spaces, and that the Shannon, Gibbs and von Neumann entropies can be recovered in the specific cases of discrete classical, continuous classical, and quantum mechanics.
	\item Show that the geometric structures of classical and quantum mechanics can be recovered from the entropic structure alone.
	\item Demonstrate that information geometry can be generalized to the ensemble space of any physical theory of finite dimensions. 
	\item Show that the entropic structure imposes constraints on the space of possible physical theories and rules out some Generalized Probabilistic Theories.
\end{itemize}
The above expected results should lead naturally to a range of additional possible directions and questions that could be addressed during or subsequent to the grant period, depending on the rate of advancement.  For example:
\begin{itemize}
	\item Identify what results from information geometry and information theory could be generalized to ensemble spaces; this would lend itself to various self-contained contributions by the grad student, undergrads, or individuals outside the project.
	\item Determine whether the convexity of the mixing entropy can be physically motivated and whether that guarantees a continuous embedding into a locally convex topological vector space.
	\item Generalize information geometry to the infinite-dimensional case; this would require a locally convex topology to be able to define variations.
	\item Determine whether the ensemble spaces can be in general proven to be metrizable and characterize the connection between the metric and the entropy.
	\item Find a suitable function space that can be used as a model for the case of classical field theories.
	\item Generalize classical Hamiltonian mechanics and the principle of stationary action to a generic ensemble space.
\end{itemize}

\textcolor{blue}{The immediate significance of the proposed research stems from the realization that, though classical information theory and quantum information theory are presented as two separate subjects in textbooks, their numerous corresponding features make them an interesting target for generalization.  The broader significance lies in the overall approach of the research as well as its specific potential to provide a common axiomatic foundation for \emph{all} physical theories.  Currently within physics, we typically do not know from where the various mathematical structures emerge, making it unclear in which directions they can be generalized in order to develop new theories that are physically meaningful.  Our approach instead seeks to \emph{find axioms from physical requirements}, i.e.~the axiomatics are justified from the physics.  Such an axiomatic foundation common to all physical theories would serve not only to better ground current theories and elucidate their relationships, but also to constrain potential future physical theories.} While there exists other work attempting to reconstruct quantum mechanics from physical principles, for example in Refs.~\cite{hardy2001quantumaxioms,caticha2008,dakic2009,Chiribella2010,Masanes_2011}, we are not aware of any literature attempting to relate the geometric and entropic structures of either classical or quantum mechanics, much less those of a general physical theory. Works trying to reconstruct quantum mechancis from physical premises are typically in finite dimensions, do not generalize to classical mechanics, and do not provide a mathematically rigorous path to generalize to classical or quantum field theories.

@@

### 2025-11-08 Proposals/2025-NSF-EntropicGeometry/ProjectSummary.tex
Final version of project summary, significance, and bibliography formatting before talking to Alejandro

\noindent
\textbf{Intellectual Merit:}  \\
\textcolor{blue}{The primary advancement will be to identify and characterize a set of mathematical structures common to \emph{any} physical theory, explicitly showing how these general structures reduce to the standard structures in the cases of discrete classical, continuous classical, and quantum systems. Such a generalization will provide further insight into the relationships among established physical theories and place constraints on the allowed mathematical structures in potential future physical theories. All axioms will be physically motivated and stated as properties of ensemble spaces.}

\textcolor{blue}{The proposed work starts from the observation that the spaces of statistical ensembles of classical and quantum mechanics have multiple features in common. They are both convex subsets of a locally convex topological vector space whose metric is induced by the information entropy, or more specifically, by the Jensen-Shannon divergence. This is in contrast to the different structures of the spaces of classical and quantum states, which rely on symplectic manifolds and projective spaces of complex Hilbert spaces, respectively. Even more notably, our preliminary work shows that the symplectic form of classical mechanics and the inner product (i.e.~Born rule) of quantum mechanics can indeed be recovered by the entropy. In general, the states defined by mechanics can be understood as special cases of ensembles with the lowest possible entropy or the extreme points of the convex set.
While classical mechanics and quantum mechanics have a different structure, classical and quantum statistical mechanics have common features.  Moreover, classical and quantum information theory have numerous corresponding features.  Our investigation will take those common features and generalize them, providing a common foundational structure to both classical and quantum theories.
}

>>>>

\noindent
\textbf{Intellectual Merit:}  \\
\textcolor{blue}{While the state spaces of classical mechanics and quantum mechanics have different structures, symplectic manifolds and projective spaces of complex Hilbert spaces, respectively, the spaces of classical and quantum \emph{statistical ensembles} have common features.  They are both convex subsets of a locally convex topological vector space whose metric is induced by the information entropy, or more specifically, by the Jensen-Shannon divergence.  Moreover, classical and quantum information theory have numerous corresponding features.  The primary advancement of the research will be to take those common and corresponding features and generalize them, providing a common, mathematically rigorous, foundational structure to both classical and quantum theories.  Such a generalization will provide further insight into the relationships among established physical theories and place constraints on the allowed mathematical structures in potential future physical theories. All axioms will be physically motivated and stated as properties of ensemble spaces.  Preliminary work shows that the symplectic form of classical mechanics and the inner product (i.e.~Born rule) of quantum mechanics can be recovered by the entropy, tightly linking the geometric and entropic structures, and that the states defined by mechanics can be understood as special cases of ensembles with the lowest possible entropy or the extreme points of the convex set. }

\textcolor{blue}{
}
